---
title: "High-Performance Graph Analysis with graphfast"
author: "Kyle Haynes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{High-Performance Graph Analysis with graphfast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

# Introduction

The `graphfast` package provides high-performance algorithms for analyzing large-scale graphs in R. Built with optimized C++ implementations, it can handle graphs with hundreds of millions of edges efficiently. The package also includes graph-based entity resolution for deduplication and record linkage.

## Key Features

- **Graph Analysis**: Connected components, shortest paths, connectivity queries
- **Entity Resolution**: Multi-column grouping with the `group_id()` function
- **Memory Efficiency**: Sparse representations and streaming algorithms
- **Performance**: Optimized C++ implementations for massive datasets

## Installation

```{r eval=FALSE}
# From GitHub
devtools::install_github("KyleHaynes/graphfast")

# Build from source
devtools::install()
```

```{r}
library(graphfast)
```

## Entity Resolution with group_id()

The `group_id()` function provides high-performance entity resolution and deduplication across multiple columns. It's perfect for finding records that represent the same entity despite variations in contact information.

### Basic Entity Resolution

```{r}
# Customer data with potential duplicates
customers <- data.frame(
  customer_id = 1:6,
  name = c("John Smith", "Jane Doe", "J. Smith", "Bob Wilson", "Alice Brown", "John S."),
  phone1 = c("123-456-7890", "987-654-3210", "123-456-7890", "", "555-0123", ""),
  phone2 = c("", "987-654-3210", "555-1234", "123-456-7890", "", "123-456-7890"),
  email = c("john@email.com", "jane@email.com", "john2@email.com", "john@email.com", "alice@email.com", ""),
  stringsAsFactors = FALSE
)

print(customers)
```

```{r}
# Find groups based on shared phone numbers and emails
group_ids <- group_id(customers, 
                      cols = c("phone1", "phone2", "email"),
                      incomparables = c("", "NA"))

customers$group_id <- group_ids
print(customers)
```

### Advanced Entity Resolution

```{r}
# Get detailed results with value mappings
result <- group_id(customers, 
                   cols = c("phone1", "phone2", "email"),
                   incomparables = c(""),
                   return_details = TRUE)

print(result)
```

### Regex Column Selection

The `group_id()` function supports regex patterns for column selection:

```{r}
# Use regex to match column names (default behavior)
group_ids_regex <- group_id(customers, 
                           cols = "phone",  # Matches "phone1", "phone2"
                           incomparables = c(""))

# Use exact column names
group_ids_exact <- group_id(customers, 
                           cols = c("phone1", "phone2"),
                           use_regex = FALSE,
                           incomparables = c(""))

print("Results are identical:", identical(group_ids_regex, group_ids_exact))
```

## Graph Analysis

### Creating a Graph

Graphs in `graphfast` are represented as edge lists - two-column matrices where each row represents an edge between two nodes.

```{r}
# Create a simple graph
edges <- matrix(c(
  1, 2,
  2, 3,
  3, 4,
  1, 4,  # Creates a cycle
  5, 6,  # Separate component
  7, 8,
  8, 9
), ncol = 2, byrow = TRUE)

print(edges)
```

### Finding Connected Components

Connected components are maximal sets of nodes where every pair is connected by a path.

```{r}
components <- find_connected_components(edges)
print(components)
```

The result contains:
- `components`: Vector assigning each node to a component ID
- `component_sizes`: Size of each component  
- `n_components`: Total number of components

### Checking Connectivity

You can efficiently check if specific pairs of nodes are connected:

```{r}
# Check if these pairs are connected
query_pairs <- matrix(c(
  1, 3,  # Same component
  1, 5,  # Different components  
  7, 9   # Same component
), ncol = 2, byrow = TRUE)

connected <- are_connected(edges, query_pairs)
print(connected)
```

### Computing Shortest Paths

Find shortest path distances between node pairs:

```{r}
distances <- shortest_paths(edges, query_pairs)
print(distances)
```

A distance of -1 indicates no path exists between the nodes.

## Performance Comparison

Let's compare the performance of different approaches:

### Entity Resolution Performance

```{r}
# Generate larger dataset for entity resolution
set.seed(123)
n_records <- 1000

large_customers <- data.frame(
  id = 1:n_records,
  phone1 = sample(c(paste0("555-", sample(1000:9999, 200)), rep("", 50)), n_records, replace = TRUE),
  phone2 = sample(c(paste0("555-", sample(1000:9999, 200)), rep("", 50)), n_records, replace = TRUE),
  email = sample(c(paste0("user", 1:300, "@email.com"), rep("", 50)), n_records, replace = TRUE),
  stringsAsFactors = FALSE
)

# Benchmark entity resolution
entity_time <- system.time({
  large_groups <- group_id(large_customers, 
                          cols = c("phone1", "phone2", "email"),
                          incomparables = c(""))
})

cat("Entity resolution time:", entity_time["elapsed"], "seconds\n")
cat("Records processed:", n_records, "\n")
cat("Groups found:", max(large_groups), "\n")
cat("Throughput:", round(n_records / entity_time["elapsed"]), "records/second\n")
```

### Graph Analysis Performance

Let's test performance on a larger graph:

```{r}
# Generate a random graph with 10,000 nodes and 50,000 edges
set.seed(123)
n_nodes <- 10000
n_edges <- 50000

large_edges <- matrix(
  sample(1:n_nodes, 2 * n_edges, replace = TRUE), 
  ncol = 2
)

# Remove self-loops
large_edges <- large_edges[large_edges[,1] != large_edges[,2], ]

cat("Graph size:", nrow(large_edges), "edges,", n_nodes, "nodes\n")
```

```{r}
# Benchmark connected components
system.time({
  large_result <- find_connected_components(large_edges)
})

cat("Found", large_result$n_components, "components\n")
cat("Largest component:", max(large_result$component_sizes), "nodes\n")
```

### Memory Efficiency

The package uses memory-efficient algorithms:

```{r}
# Graph statistics without storing full adjacency matrix
stats <- graph_statistics(large_edges, n_nodes = n_nodes)
print(stats)
```

## Real-World Applications

### Customer Deduplication

```{r}
# Realistic customer deduplication scenario
customers_raw <- data.frame(
  customer_id = 1:10,
  name = c("John Smith", "Jane Doe", "J. Smith", "Bob Wilson", "Alice Brown", 
           "John S.", "Jane D.", "Robert W.", "John Smith", "B. Wilson"),
  phone1 = c("123-456-7890", "987-654-3210", "123-456-7890", "555-0123", "555-9876",
             "", "987-654-3210", "555-0123", "123-456-7890", ""),
  phone2 = c("", "987-654-3210", "555-1234", "123-456-7890", "",
             "123-456-7890", "", "999-888-7777", "", "555-0123"),
  email = c("john@email.com", "jane@email.com", "john2@email.com", "bob@email.com", "alice@email.com",
            "", "jane@email.com", "rob@email.com", "john@email.com", "bob@email.com"),
  stringsAsFactors = FALSE
)

# Perform deduplication
dedup_groups <- group_id(customers_raw, 
                        cols = c("phone1", "phone2", "email"),
                        incomparables = c("", "NA", "Unknown"),
                        min_group_size = 1)

customers_raw$duplicate_group <- dedup_groups

# Show results
print(customers_raw)

# Summary of duplicates
duplicate_summary <- table(customers_raw$duplicate_group)
cat("Duplicate groups found:", sum(duplicate_summary > 1), "\n")
cat("Total duplicates identified:", sum(duplicate_summary) - sum(duplicate_summary == 1), "\n")
```

### Social Network Analysis

```{r}
# Simulate a social network with communities
generate_social_network <- function(n_users, connection_prob) {
  edges <- matrix(nrow = 0, ncol = 2)
  
  for (i in 1:(n_users-1)) {
    for (j in (i+1):n_users) {
      if (runif(1) < connection_prob) {
        edges <- rbind(edges, c(i, j))
      }
    }
  }
  
  return(edges)
}

# Generate a small social network
social_edges <- generate_social_network(100, 0.05)
cat("Social network with", nrow(social_edges), "friendships\n")

# Find communities (connected components)
communities <- find_connected_components(social_edges, n_nodes = 100)
cat("Found", communities$n_components, "communities\n")
cat("Community sizes:", paste(sort(communities$component_sizes, decreasing = TRUE), collapse = ", "), "\n")
```

### Web Graph Analysis

```{r}
# Simulate web page links
web_pages <- 500
link_prob <- 0.01

web_edges <- matrix(nrow = 0, ncol = 2)
for (i in 1:web_pages) {
  # Each page links to a few others
  n_links <- rpois(1, 3)
  if (n_links > 0) {
    targets <- sample(setdiff(1:web_pages, i), min(n_links, web_pages-1))
    for (target in targets) {
      web_edges <- rbind(web_edges, c(i, target))
    }
  }
}

cat("Web graph with", nrow(web_edges), "links between", web_pages, "pages\n")

# Analyze link structure  
web_stats <- graph_statistics(web_edges, n_nodes = web_pages)
cat("Web graph density:", round(web_stats$density, 4), "\n")
cat("Average out-degree:", round(web_stats$degree_stats$mean, 2), "\n")

# Find strongly connected components
web_components <- find_connected_components(web_edges, n_nodes = web_pages)
cat("Found", web_components$n_components, "connected regions\n")
```

## Advanced Features

### Batch Processing

For maximum efficiency, process multiple queries together:

```{r}
# Generate many connectivity queries
n_queries <- 1000
query_nodes <- sample(1:100, 2 * n_queries, replace = TRUE)
batch_queries <- matrix(query_nodes, ncol = 2)

# Process all queries at once
system.time({
  batch_results <- are_connected(social_edges, batch_queries, n_nodes = 100)
})

cat("Processed", n_queries, "queries\n")
cat("Connected pairs:", sum(batch_results), "out of", n_queries, "\n")
```

### Memory Monitoring

```{r}
# Monitor memory usage for large graphs
monitor_memory <- function(expr) {
  gc_before <- gc()
  result <- expr
  gc_after <- gc()
  
  memory_diff <- (gc_after["Vcells", "used"] - gc_before["Vcells", "used"]) * 8 / 1024^2
  cat("Memory used:", round(memory_diff, 2), "MB\n")
  
  return(result)
}

# Test memory usage
monitor_memory({
  test_result <- find_connected_components(large_edges)
})
```

## Algorithm Details

### Connected Components
- Uses Union-Find with path compression and union by rank
- Time complexity: O(m × α(n)) where α is inverse Ackermann function  
- Space complexity: O(n)

### Entity Resolution (group_id)
- Union-Find algorithm for optimal grouping performance
- Multi-column matching with configurable incomparable values
- Time complexity: O(m × α(n)) where m is total values across columns
- Memory efficient: processes millions of records with minimal overhead

### Shortest Paths
- Breadth-first search from each source node
- Time complexity: O(m + n) per query
- Early termination with maximum distance limits

### Memory Optimization
- Sparse graph representation using adjacency lists
- No full adjacency matrix storage
- Streaming algorithms for very large datasets

## Best Practices

### Graph Analysis
1. **Use integer node IDs** starting from 1 for best performance
2. **Specify n_nodes** when known to avoid scanning edges
3. **Batch queries** together rather than processing individually  
4. **Set distance limits** for shortest path queries when appropriate
5. **Monitor memory** usage for very large graphs

### Entity Resolution
1. **Specify incomparable values** to exclude empty strings, NAs, etc.
2. **Use regex column selection** for flexible column matching
3. **Set minimum group sizes** to filter out singleton matches
4. **Consider case sensitivity** based on your data characteristics
5. **Use return_details=TRUE** for debugging and analysis

## Conclusion

The `graphfast` package provides a focused suite of high-performance tools for:

- **Graph Analysis**: Connected components, shortest paths, and connectivity queries for massive graphs
- **Entity Resolution**: Multi-column deduplication and record linkage with the `group_id()` function

Its optimized C++ implementations enable analysis of datasets that would be intractable with traditional R approaches, making it suitable for big data applications in:

- Social network analysis and community detection
- Customer deduplication and master data management
- Bioinformatics and protein interaction networks
- Web graph analysis and link structure
- Infrastructure network analysis

The package scales efficiently to handle millions of records and hundreds of millions of edges while maintaining memory efficiency and providing intuitive R interfaces.