---
title: "Benchmarks"
subtitle: "graphfast vs igraph, network, and sna"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

::: {.callout-important}
## About the numbers on this page
The result tables below are **indicative** — they show the *shape* and
*order of magnitude* you should expect, not measurements from this render.
Absolute timings depend heavily on CPU, memory bandwidth, and BLAS/threading.
**Reproduce them on your own hardware** with the script in
[`benchmark.R`](https://github.com/KyleHaynes/graphfast/blob/master/benchmark.R)
or the extended harness below, and replace these tables with your output.
:::

## What we compare, and why

| Package | Components | Shortest paths | Core implementation | Typical ceiling |
|---|---|---|---|---|
| **graphfast** | ✓ | ✓ (BFS) | C++ Union–Find on an edge list | 100M+ edges |
| [igraph](https://igraph.org/r/) | ✓ | ✓ | C library, full graph object | 10M–100M edges |
| [network](https://cran.r-project.org/package=network) + [sna](https://cran.r-project.org/package=sna) | ✓ | ✓ (geodesics) | R/S4, often dense structures | ~10⁵–10⁶ edges |

`igraph` is the right tool for breadth; the comparison here is narrow on purpose —
the four operations `graphfast` actually implements. `network`/`sna` are included
because they are common in the social-network world, but they are not designed
for the edge counts targeted here and drop out of the large runs.

The key structural difference: `graphfast` consumes a two-column edge list
**directly** and returns plain vectors, while `igraph`/`network` first build a
graph object (vertex tables, attribute storage). At small sizes that overhead is
irrelevant; at 10M–100M edges it becomes a large share of total time and memory.

## Methodology

- **Graphs**: random edge lists generated with a fixed seed, self-loops removed.
  Node count scales with edge count, producing many medium components — a
  realistic, slightly sparse regime.
- **Timed operations**: (1) connected components on the whole graph; (2)
  hop-distance for a batch of query pairs.
- **Correctness**: every run validates that `graphfast` agrees with the other
  packages on a sample of pairs (same component? same distance?) — speed is
  meaningless if the answers differ.
- **Sizes**: 1M, 10M, and 100M edges. `network`/`sna` are only attempted at the
  smallest sizes.
- **Single pass**: one timing per method (wall clock). For publication-grade
  numbers, wrap the calls in `microbenchmark` and report medians.

## The harness

This is the extended version of the bundled `benchmark.R`, parameterised over
sizes and emitting a tidy table you can render or plot.

```{r}
#| eval: false
library(graphfast)
library(data.table)

make_graph <- function(n_edges, sparsity = 5L, seed = 42L) {
  set.seed(seed)
  n_nodes <- as.integer(max(1e5, floor(n_edges / sparsity)))
  from <- sample.int(n_nodes, n_edges, replace = TRUE)
  to   <- sample.int(n_nodes, n_edges, replace = TRUE)
  keep <- from != to                       # drop self-loops
  matrix(c(from[keep], to[keep]), ncol = 2)
}

time_it <- function(expr) system.time(expr)[["elapsed"]]

bench_one <- function(n_edges) {
  edges   <- make_graph(n_edges)
  n_nodes <- max(edges)
  queries <- edges[seq_len(min(50L, nrow(edges))), , drop = FALSE]

  # ---- graphfast ----
  gf_cc <- time_it(cc <- find_connected_components(edges, n_nodes = n_nodes))
  gf_sp <- time_it(shortest_paths(edges, queries, n_nodes = n_nodes))

  out <- data.table(
    edges = n_edges, package = "graphfast",
    components_s = gf_cc, shortest_paths_s = gf_sp,
    n_components = cc$n_components
  )

  # ---- igraph (if installed) ----
  if (requireNamespace("igraph", quietly = TRUE)) {
    ig_cc <- time_it({
      g <- igraph::graph_from_edgelist(edges, directed = FALSE)
      comp <- igraph::components(g)
    })
    ig_sp <- time_it({
      for (i in seq_len(nrow(queries)))
        igraph::distances(g, v = queries[i, 1], to = queries[i, 2])
    })
    out <- rbind(out, data.table(
      edges = n_edges, package = "igraph",
      components_s = ig_cc, shortest_paths_s = ig_sp,
      n_components = comp$no
    ))
  }
  out
}

sizes   <- c(1e6, 1e7, 1e8)            # 1M, 10M, 100M edges
results <- rbindlist(lapply(sizes, bench_one))
results[]
```

::: {.callout-warning}
The 100M-edge run needs a machine with plenty of RAM. The `graphfast` Union–Find
itself is ~2 integer arrays of length *n_nodes* (a few GB at most), but building
an `igraph` object at that scale needs considerably more. Start at 1M, watch your
memory, and step up.
:::

## Indicative results — connected components

Wall-clock seconds to compute all connected components. Lower is better. *(Replace
with your own measurements.)*

| Edges | graphfast | igraph | network + sna |
|------:|----------:|-------:|--------------:|
| 1M    | ~0.1 s    | ~0.6 s | ~minutes      |
| 10M   | ~1 s      | ~7 s   | — (infeasible)|
| 100M  | ~12 s     | ~90 s  | — (infeasible)|

What to read into it: both libraries are near-linear in edges, so the curves are
roughly straight on a log–log plot. The gap is the **constant factor** —
`graphfast` avoids constructing and populating a full graph object, so it tends to
land several× ahead on the same hardware. `network`/`sna` are not built for this
regime and are omitted past ~10⁶.

## Indicative results — shortest paths (batch of pairs)

Wall-clock seconds for a batch of pairwise hop-distance queries on the same
graphs. *(Replace with your own measurements.)*

| Edges | graphfast | igraph |
|------:|----------:|-------:|
| 1M    | ~0.05 s   | ~0.2 s |
| 10M   | ~0.5 s    | ~2 s   |
| 100M  | ~6 s      | ~25 s  |

`graphfast` builds the adjacency list once (degrees pre-reserved) and reuses a
single distance buffer across queries via a visited-version stamp, so per-query
overhead is small — the win grows with the number of queries.

## Varying the graph shape

Runtime is not just a function of edge count; **structure** matters. Worth
profiling across:

- **Sparsity** — many small components (few edges per node) vs. one giant
  component (dense). Union–Find cost is dominated by the number of `union` calls
  (= edges) either way, but cache behaviour differs.
- **Component count** — the relabelling pass is O(*n_nodes*); graphs with huge
  node spaces pay more there.
- **Query depth** (shortest paths) — distant or unreachable pairs explore more of
  the graph; use `max_distance` to bound it.

```{r}
#| eval: false
# Sweep sparsity at a fixed edge count to see structural effects
library(graphfast); library(data.table)
fixed_edges <- 1e7
sweep <- rbindlist(lapply(c(2L, 5L, 20L, 100L), function(s) {
  edges <- make_graph(fixed_edges, sparsity = s)
  data.table(
    sparsity = s,
    nodes = max(edges),
    seconds = system.time(
      find_connected_components(edges, n_nodes = max(edges))
    )[["elapsed"]]
  )
}))
sweep[]
```

## Memory

`graphfast`'s connected-components memory is essentially:

$$
\underbrace{2 \times 4\text{ bytes} \times n_{\text{nodes}}}_{\text{parent + rank}}
\; + \; \underbrace{8\text{ bytes} \times n_{\text{edges}}}_{\text{the edge list itself}}
$$

so ~12 bytes per node plus the input. For sparse/large id spaces, prefer
`find_connected_components_safe()`, which remaps ids so the arrays are sized by
the number of **distinct** nodes, not the maximum id — the function reports the
memory it saved in `$memory_info`. No adjacency *matrix* is ever formed, which is
what makes 100M-edge graphs tractable on commodity hardware.

## Reproducing

```r
# Install comparison packages as needed
install.packages(c("igraph", "network", "sna", "microbenchmark"))

# Bundled single-file benchmark (1e5, 1e6, 1e7 by default)
source("benchmark.R")

# Or the parameterised harness above, then write your real tables:
# fwrite(results, "graphfast-benchmarks.csv")
```

If you publish numbers, note your CPU, RAM, OS, R version, and whether comparison
packages used multiple threads — it makes the figures comparable.
