---
title: "Examples"
subtitle: "Every function, demonstrated"
---

```{r}
#| label: setup
#| message: false
#| warning: false
# Install graphfast from GitHub if it isn't already available...
# if (!requireNamespace("graphfast", quietly = TRUE)) {
#   if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
#   remotes::install_github("KyleHaynes/graphfast")
# }

# # ...along with the optional (Suggests) packages these examples use.
# for (pkg in c("data.table", "visNetwork")) {
#   if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
# }

print(.libPaths())
library(graphfast)
library(data.table)
set.seed(1)
```

This page walks through every exported function with small, self-contained
examples. The chunks are live — the output you see is produced when the site is
rendered.

A graph in `graphfast` is just a **two-column edge list**: a matrix or
`data.table` where each row is an edge `from -> to`. We'll reuse this little
graph with three components throughout:

```{r}
#| label: demo-graph
edges <- matrix(c(
  1, 2,
  2, 3,
  3, 4,   # component A: {1,2,3,4}
  5, 6,   # component B: {5,6}
  7, 8,
  8, 9    # component C: {7,8,9}
), ncol = 2, byrow = TRUE)
edges
```

# Connected components

## `find_connected_components()`

The workhorse. Returns one component id **per node**, the size of each component,
and the total count.

```{r}
cc <- find_connected_components(edges)
cc$n_components
cc$component_sizes
cc$components          # component id for node 1, 2, 3, ...
```

Node 5 and node 6 share component 2; node 1 is in component 1:

```{r}
cc$components[c(1, 5, 6)]
```

`compress = FALSE` keeps the raw Union–Find root labels instead of tidy
consecutive ids (the counts are still correct):

```{r}
raw <- find_connected_components(edges, compress = FALSE)
raw$n_components
head(raw$components)
```

You can pass `n_nodes` explicitly to skip the scan for the maximum id (handy when
you already know it, or to include isolated nodes with no edges):

```{r}
find_connected_components(edges, n_nodes = 12)$n_components   # nodes 10,11,12 are singletons
```

## `find_connected_components_safe()`

When node ids are **sparse or large**, a dense `parent` array up to the maximum
id would waste memory. This version remaps ids to `1..(unique nodes)` with
`fastmatch`, runs the algorithm, then maps the answer back. The returned
`components` vector is **named by the original ids**.

```{r}
sparse_edges <- matrix(c(
  1000, 2000,
  2000, 3000,
  9000, 9001
), ncol = 2, byrow = TRUE)

res <- find_connected_components_safe(sparse_edges, verbose = FALSE)
res$n_components
res$components            # names are the original node ids
res$memory_info$memory_saved_gb
```

## `find_connected_components_large()`

Handles ids **beyond the 32-bit integer limit** (≈ 2.1e9) by keeping them as
doubles until after remapping. Set `verbose = TRUE` to see what it decided.

```{r}
big_edges <- matrix(c(
  1e10, 2e10,
  2e10, 3e10,
  7e10, 8e10
), ncol = 2, byrow = TRUE)

big <- find_connected_components_large(big_edges, verbose = FALSE)
big$n_components
big$components
```

# Per-edge components

Often you don't want a node-indexed vector — you want a component id attached to
each **edge / row** of your data.

## `get_edge_components()`

```{r}
ec <- get_edge_components(edges)
ec$from_components       # component of each edge's 'from' node
ec$to_components         # ... and its 'to' node (always equal for a real edge)
ec$n_components
```

`return_type = "combined"` returns just the `from_components` vector (same length
as the number of edges):

```{r}
get_edge_components(edges, return_type = "combined")
```

## `group_edges()` and `edge_components()`

`group_edges()` is the convenience alias returning one component id per edge row.
`edge_components()` does the same but takes a `data.table` and column names, so it
drops straight into a `data.table` expression.

```{r}
# Matrix in -> vector out
group_edges(edges)
```

```{r}
dt <- data.table(from = c(1, 2, 5), to = c(2, 3, 6))
dt[, component := edge_components(.SD, "from", "to")]
dt
```

## `add_component_column()`

Adds a component column to a `data.table` **by reference** (no copy), mapping node
ids with `fastmatch` internally. Works with any column names.

```{r}
dt2 <- data.table(source = c(1, 2, 5, 7), target = c(2, 3, 6, 8), weight = runif(4))
add_component_column(dt2, from_col = "source", to_col = "target")
dt2
```

Use `in_place = FALSE` to leave the original untouched and get a copy back, and
`verbose = TRUE` for per-step timing on large data:

```{r}
out <- add_component_column(dt2, from_col = "source", to_col = "target",
                            component_col = "grp", in_place = FALSE)
out
```

# Connectivity and distance

## `are_connected()`

Batched "are these two nodes in the same component?" queries. Builds Union–Find
once, then answers each pair in near-constant time.

```{r}
queries <- matrix(c(
  1, 4,   # same component A
  1, 5,   # A vs B
  7, 9    # same component C
), ncol = 2, byrow = TRUE)

are_connected(edges, queries)
```

## `shortest_paths()`

Hop-distance (unweighted BFS) for each query pair. `-1` means no path exists.

```{r}
shortest_paths(edges, queries)     # 1->4 is 3 hops; 1->5 unreachable; 7->9 is 2 hops
```

`max_distance` stops the search early — anything farther returns `-1`:

```{r}
shortest_paths(edges, matrix(c(1, 4), ncol = 2), max_distance = 2)   # 1->4 needs 3 hops
```

## `graph_statistics()`

Degree summary and density in a single pass — no adjacency matrix.

```{r}
stats <- graph_statistics(edges)
stats$n_nodes
stats$n_edges
stats$density
stats$degree_stats
```

# Entity resolution / grouping

These solve "group rows that share a value across any of these columns" as a
connected-components problem.

## `group_id()`

The general engine. Give it a `data.frame`/`data.table` and the columns to match
on (regex by default), plus values that should never glue rows together
(`incomparables`).

```{r}
customers <- data.frame(
  id     = 1:6,
  phone1 = c("123-456-7890", "987-654-3210", "123-456-7890", "",            "555-0123", ""),
  phone2 = c("",             "987-654-3210", "555-1234",     "123-456-7890","",         "555-0123"),
  email  = c("a@x.com",      "b@x.com",      "c@x.com",      "a@x.com",     "d@x.com",  "d@x.com"),
  stringsAsFactors = FALSE
)

# cols = "phone" matches phone1 + phone2 by regex; "" is incomparable
group_id(customers, cols = "phone", incomparables = c(""))
```

Match across all three columns by passing exact names (`use_regex = FALSE`):

```{r}
group_id(customers, cols = c("phone1", "phone2", "email"),
         use_regex = FALSE, incomparables = c(""))
```

`return_details = TRUE` returns a rich object with a `print` method showing which
shared values created which groups:

```{r}
res <- group_id(customers, cols = c("phone1", "phone2", "email"),
                use_regex = FALSE, incomparables = c(""), return_details = TRUE)
res
```

Other useful arguments:

```{r}
# Only keep groups of at least 2 (singletons get id 0)
group_id(customers, cols = "phone", incomparables = c(""), min_group_size = 2)

# Case-insensitive matching
group_id(data.frame(a = c("ABC", "abc", "xyz")), case_sensitive = FALSE)
```

## `add_group_ids()`

The `data.table` wrapper for `group_id()` — adds the group column by reference.

```{r}
dt_cust <- as.data.table(customers)
add_group_ids(dt_cust, cols = c("phone1", "phone2", "email"),
              group_col = "entity_id", use_regex = FALSE, incomparables = c(""))
dt_cust
```

## `set_group_id()`

An alternative grouping engine using the **bipartite edge reduction** (melt →
edges → components). It requires an `id` column and writes the group id back by
reference.

### Basic grouping

Two rows land in the same group when they share **any** value across the chosen
columns. Here rows 1 and 3 share phone `"123"`, and rows 1 and 4 share phone
`"111"`, so all three collapse into one group.

```{r}
dt_sg <- data.table(
  id     = 1:5,
  phone1 = c("123", "456", "123", "789", "456"),
  phone2 = c("111", "222", "333", "111", "222")
)
set_group_id(dt_sg, cols = "phone", var_output_name = "gid")
dt_sg
```

### Choosing columns with a regex

`cols` is a regular expression matched against the column names, so one pattern
can pull in several value columns at once. Here `"phone|email"` grabs all three.

```{r}
dt_re <- data.table(
  id     = 1:5,
  phone1 = c("123-456-7890", "987-654-3210", "123-456-7890", "", "555-0123"),
  phone2 = c("", "987-654-3210", "555-1234", "123-456-7890", ""),
  email  = c("john@email.com", "jane@email.com", "bob@email.com",
             "john@email.com", "alice@email.com")
)
set_group_id(dt_re, cols = "phone|email")
dt_re
```

### Naming the output column

`var_output_name` controls the column written back (default `"gid"`).

```{r}
dt_name <- copy(dt_re)[, gid := NULL]
set_group_id(dt_name, cols = "phone", var_output_name = "household")
dt_name
```

### Values that should never link rows

`incomparables` lists values that are ignored when forming groups — empty
strings and `NA` by default, so blank cells never tie records together. Add your
own sentinels (e.g. `"Unknown"`) as needed.

```{r}
dt_inc <- data.table(
  id    = 1:4,
  phone = c("Unknown", "Unknown", "555-1234", "555-1234"),
  alt   = c("", "", "", "555-9999")
)
# Without excluding "Unknown", rows 1 and 2 would be linked by it.
set_group_id(dt_inc, cols = "phone|alt",
             var_output_name = "gid",
             incomparables = c(NA, "", "Unknown"))
dt_inc
```

### Treating column sets in isolation

By default every selected column shares one value namespace, so a phone number
appearing in an email column would link the two. Pass `isolate` — a list of
column sets — to keep each set's values separate: identical text only links rows
**within the same set**. The columns to group on come from the union of the
sets, and `cols` is ignored.

Note row 5: its phone number `"123-456-7890"` sits in the `email` column. Because
`email` is isolated from the phone columns, that value does **not** pull row 5
into the phone group.

```{r}
dt_iso <- data.table(
  id     = 1:5,
  phone1 = c("123-456-7890", "987-654-3210", "123-456-7890", "", "555-0123"),
  phone2 = c("", "987-654-3210", "555-1234", "123-456-7890", ""),
  email  = c("john@email.com", "jane@email.com", "bob@email.com",
             "john@email.com", "123-456-7890")
)
set_group_id(dt_iso, isolate = list(c("phone1", "phone2"), "email"))
dt_iso
```

For comparison, the same data **without** isolation lets the stray phone number
in `email` link row 5 into the phone group:

```{r}
dt_noiso <- copy(dt_iso)[, gid := NULL]
set_group_id(dt_noiso, cols = "phone|email")
dt_noiso
```

### Returning the edge list of pair connections

Pass `return_edges = TRUE` to get back the **record-to-record edge list** — every
pair of `id`s that share a value, with the shared `value` itself. The group-id
column is still written to `dt` by reference, so you get both the grouping *and*
the underlying connections in one call.

```{r}
dt_edges <- copy(dt_iso)[, gid := NULL]
edges <- set_group_id(dt_edges, isolate = list(c("phone1", "phone2"), "email"),
                      return_edges = TRUE)
edges          # from / to / value
dt_edges       # gid still added by reference
```

### Plotting the linkage with `visNetwork`

The edge list drops straight into [`visNetwork`](https://datastorm-open.github.io/visNetwork/):
use the `id`/`gid` columns for the nodes (colouring by group) and the returned
edges for the connections. The shared value becomes the edge tooltip.

```{r}
#| eval: !expr requireNamespace("visNetwork", quietly = TRUE)
library(visNetwork)

nodes <- data.frame(
  id    = dt_edges$id,
  label = paste0("rec ", dt_edges$id),
  group = dt_edges$gid
)
vis_edges <- data.frame(
  from  = edges$from,
  to    = edges$to,
  title = edges$value   # hover shows the value that links the pair
)

visNetwork(nodes, vis_edges, width = "100%", height = "320px") |>
  visNodes(shape = "dot", size = 22) |>
  visOptions(highlightNearest = TRUE) |>
  visLegend()
```

```{r}
#| eval: !expr requireNamespace("visNetwork", quietly = TRUE)
library(visNetwork)

nodes <- data.table(
  id = 1:5,
  phone1 = c("123-456-7890", "987-654-3210", "123-456-7890", "", "555-0123"),
  phone2 = c("", "987-654-3210", "555-1234", "123-456-7890", ""),
  email = c("john@email.com", "jane@email.com", "bob@email.com",
            "john@email.com", "123-456-7890")
)
nodes[, title := paste0("phone1: ", phone1, "</br>phone2: ", phone2, "</br>email: ", email)]

edges <- set_group_id(nodes, isolate = list(c("phone1", "phone2"), "email"),
                      return_edges = TRUE)

visNetwork(nodes, edges, width = "100%", height = "320px") |>
  visNodes(shape = "dot", size = 22) |>
  visOptions(highlightNearest = TRUE) |>
  visLegend()
```

```{r}
#| eval: !expr requireNamespace("visNetwork", quietly = TRUE)
library(visNetwork)

n_queries <- 100
query_nodes <- sample(1:100, 2 * n_queries, replace = TRUE)
batch_queries <- data.table(matrix(query_nodes, ncol = 2))
batch_queries[, id := 1:.N]
batch_queries[, title := paste0("V1: ", V1, ", V2: ", V2)]

edges <- set_group_id(batch_queries, "V", return_edges = TRUE)

visNetwork(batch_queries, edges, width = "100%", height = "320px") |>
  visNodes(shape = "dot", size = 22) |>
  visOptions(highlightNearest = TRUE) |>
  visLegend()
```


# Putting it together

A realistic `data.table` flow: take an edge list, attach components, and inspect
the largest blob — all without ever building a graph object.

```{r}
g <- data.table(
  from = sample(1:50, 200, replace = TRUE),
  to   = sample(1:50, 200, replace = TRUE)
)
g <- g[from != to]

g[, component := edge_components(.SD, "from", "to")]

# Size of each component, largest first
sizes <- g[, .N, by = component][order(-N)]
head(sizes)
```

See [Benchmarks](benchmarks.qmd) for how these scale to millions and hundreds of
millions of edges, and [Reference](reference.qmd) for the full argument lists.
