---
title: "graphfast"
subtitle: "Fast graph connectivity analysis and entity resolution in R, powered by C++"
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What is graphfast?

`graphfast` is a small, focused R package for the operations that dominate
real-world graph work on **large** data:

- **Connected components** — which nodes/edges belong to the same blob?
- **Connectivity queries** — are these two nodes reachable from each other?
- **Shortest paths** — how many hops between these node pairs?
- **Graph statistics** — degree distribution, density, counts.
- **Entity resolution** — group records that share *any* value across *any*
  of several columns (deduplication / record linkage), solved as a graph
  problem.

It is built on a single, well-understood idea — the **Union–Find** (disjoint-set)
data structure — implemented in C++ via [Rcpp](https://www.rcpp.org/) and wired
to [data.table](https://r-datatable.com/) and
[fastmatch](https://cran.r-project.org/package=fastmatch) so the R-side glue is
never the bottleneck.

::: {.callout-note}
## Scope: graphs, not everything
For comprehensive graph analysis — centrality, community detection, layouts,
isomorphism, flows — use [igraph](https://igraph.org/r/), which is mature and
feature-complete. `graphfast` deliberately trades breadth for **speed and memory
efficiency** on a handful of high-volume operations. If you have hundreds of
millions of edges and only need components / connectivity / hop-distance /
grouping, this package is built for exactly that.
:::

## Why it exists

The standard pattern in R is "build an `igraph` object, then ask a question of
it". For very large edge lists that construction step — and the vertex
bookkeeping around it — can dominate runtime and memory. `graphfast` skips the
heavyweight graph object: it consumes a **two-column edge list** (matrix or
`data.table`) directly and returns plain integer vectors, so it composes
naturally with a `data.table` pipeline:

```r
library(graphfast)
library(data.table)

dt <- data.table(from = c(1, 2, 5), to = c(2, 3, 6))

# Component id per edge, added by reference — no graph object, no copy
dt[, component := edge_components(.SD, "from", "to")]
```

## Installation

```r
# Development version from GitHub
# install.packages("remotes")
remotes::install_github("KyleHaynes/graphfast")
```

`graphfast` compiles C++ (Rcpp / C++11), so you need a working toolchain
(Rtools on Windows, Xcode CLT on macOS, `r-base-dev` on Linux).

## A 60-second tour

```{r}
#| eval: false
library(graphfast)

edges <- matrix(c(
  1, 2,
  2, 3,
  3, 4,   # component A: {1,2,3,4}
  5, 6,   # component B: {5,6}
  7, 8,
  8, 9    # component C: {7,8,9}
), ncol = 2, byrow = TRUE)

# 1. Connected components
cc <- find_connected_components(edges)
cc$n_components       #> 3
cc$component_sizes    #> 4 2 3

# 2. Are these pairs connected?
are_connected(edges, matrix(c(1, 4, 1, 5, 7, 9), ncol = 2, byrow = TRUE))
#> TRUE FALSE TRUE

# 3. Shortest path (hops); -1 means unreachable
shortest_paths(edges, matrix(c(1, 4, 1, 5), ncol = 2, byrow = TRUE))
#> 3 -1

# 4. Graph statistics
graph_statistics(edges)$density
```

## How the book is organised

| Chapter | What you'll find |
|---------|------------------|
| [Theory](theory.qmd) | The data structures and algorithms, complexity, and the design choices behind the speed. |
| [Examples](examples.qmd) | Every exported function demonstrated with small, runnable snippets. |
| [Benchmarks](benchmarks.qmd) | Live, measured comparison against `igraph` at 1M, 10M, and 100M edges. |
| [Reference](reference.qmd) | Compact API documentation — signatures, arguments, return values. |

## At a glance

| Function | Purpose |
|----------|---------|
| `find_connected_components()` | All components for a dense, small-integer node space. |
| `find_connected_components_safe()` | Same, with automatic node-ID remapping for sparse/large IDs. |
| `find_connected_components_large()` | Same, supporting IDs beyond 32-bit integers. |
| `get_edge_components()` | Component id for each *edge* (from/to). |
| `group_edges()` / `edge_components()` | One component id per edge row — ideal for `data.table`. |
| `add_component_column()` | Add a component column to a `data.table` by reference. |
| `are_connected()` | Batched pairwise connectivity queries. |
| `shortest_paths()` | Batched BFS hop-distances between node pairs. |
| `graph_statistics()` | Edges, nodes, density, degree summary. |
| `group_id()` / `add_group_ids()` / `set_group_id()` | Multi-column entity resolution via Union–Find. |

License: MIT. Source on [GitHub](https://github.com/KyleHaynes/graphfast).
