---
title: "Reference"
subtitle: "API documentation"
---

```{r}
#| include: false
knitr::opts_chunk$set(eval = FALSE)
```

Compact reference for every exported function. Conventions:

- **edges** — a two-column matrix or `data.frame`/`data.table`; each row is an
  edge `from -> to`. Node ids are positive integers.
- **query_pairs** — a two-column matrix of node pairs.
- Component / group ids are 1-based and consecutive when `compress = TRUE`
  (the default). `0` denotes "no group" where applicable.

---

# Connected components

## find_connected_components

```{r}
find_connected_components(edges, n_nodes = NULL, compress = TRUE)
```

All connected components for a dense, small-integer node space.

| Argument | Description |
|---|---|
| `edges` | Two-column edge list. |
| `n_nodes` | Total node count. If `NULL`, inferred as `max(edges)`. Set it to include isolated nodes or to skip the max scan. |
| `compress` | `TRUE`: tidy ids `1..k`. `FALSE`: raw Union–Find root labels. |

**Returns** a list: `components` (component id per node, length `n_nodes`),
`component_sizes` (size of each component), `n_components` (count).

**Notes** — errors if node ids exceed the 32-bit integer limit or if the implied
allocation would exceed 32 GB; warns above ~8 GB. For sparse/huge ids use
`find_connected_components_safe()`.

```{r}
cc <- find_connected_components(matrix(c(1,2, 2,3, 5,6), ncol = 2, byrow = TRUE))
cc$n_components   # 2
```

## `find_connected_components_safe`

```{r}
find_connected_components_safe(edges, compress = TRUE, verbose = TRUE)
```

Same result, but **remaps node ids** to a dense `1..(unique nodes)` space with
`fastmatch` first — safe and memory-efficient for sparse or large ids.

| Argument | Description |
|---|---|
| `edges` | Two-column edge list (kept numeric to tolerate large ids). |
| `compress` | As above. |
| `verbose` | Print a memory analysis. Default `TRUE`. |

**Returns** `components` (named by original node ids), `component_sizes`,
`n_components`, `node_mapping` (original ↔ mapped), and `memory_info`
(`naive_memory_gb`, `efficient_memory_gb`, `memory_saved_gb`).

## `find_connected_components_large(edges, compress = TRUE, verbose = FALSE)`

For node ids **beyond the 32-bit integer range**. Ids are held as doubles and
remapped to consecutive integers internally.

**Returns** `components` (named by original ids), `component_sizes`,
`n_components`, and (when remapping occurs) `node_mapping`.

---

# Per-edge components

## `get_edge_components`

```{r}
get_edge_components(edges, n_nodes = NULL, compress = TRUE, return_type = "list")
```

Component assignment for each **edge** rather than each node.

| Argument | Description |
|---|---|
| `return_type` | `"list"`: `from_components`, `to_components`, `n_components`. `"combined"`: just the `from_components` vector. |

For a genuine edge, `from` and `to` are always in the same component, so the
combined vector is the natural "component id per row".

## `group_edges(edges, n_nodes = NULL, compress = TRUE)`

Convenience alias returning one component id per edge row
(`get_edge_components(..., return_type = "combined")`).

```{r}
group_edges(matrix(c(1,2, 2,3, 5,6), ncol = 2, byrow = TRUE))   # 1 1 2
```

## `edge_components`

```{r}
edge_components(dt, from_col, to_col, compress = TRUE)
```

Same as `group_edges()` but takes a `data.table`/`data.frame` and column names —
designed for inline `data.table` use.

```{r}
dt[, component := edge_components(.SD, "from", "to")]
```

## `add_component_column`

```{r}
add_component_column(dt, from_col = "from", to_col = "to", component_col = "component", n_nodes = NULL, compress = TRUE, in_place = TRUE, verbose = FALSE)
```

Add a component column to a `data.table`. Node ids are mapped with `fastmatch`.

| Argument | Description |
|---|---|
| `from_col`, `to_col` | Edge endpoint columns. |
| `component_col` | Name of the new column. |
| `in_place` | `TRUE`: modify `dt` by reference. `FALSE`: return a modified copy. |
| `verbose` | Print per-step timing. |

**Returns** the `data.table` (invisibly when `in_place = TRUE`).

---

# Connectivity, distance, statistics

## `are_connected`

```{r}
are_connected(edges, query_pairs, n_nodes = NULL)
```

Are the two nodes in each pair in the same component? Builds Union–Find once.

**Returns** a logical vector, one element per query pair.

```{r}
are_connected(edges, matrix(c(1,3, 1,5), ncol = 2, byrow = TRUE))   # TRUE FALSE
```

## `shortest_paths`

```{r}
shortest_paths(edges, query_pairs, n_nodes = NULL, max_distance = -1)
```

Unweighted shortest **hop** distance per query pair, via BFS.

| Argument | Description |
|---|---|
| `max_distance` | Stop searching beyond this depth; farther pairs return `-1`. `-1` (default) = no limit. |

**Returns** an integer vector; `0` for identical source/target, `-1` if no path
(or beyond `max_distance`).

## `graph_statistics`

```{r}
graph_statistics(edges, n_nodes = NULL)
```

Single-pass degree and density summary; no adjacency matrix.

**Returns** `n_edges`, `n_nodes`, `density`, and `degree_stats`
(`min`, `max`, `mean`).

---

# Entity resolution / grouping

## `group_id`

```{r}
group_id(data, cols = NULL, use_regex = TRUE, incomparables = c("", "NA", "Unknown"), case_sensitive = TRUE, min_group_size = 1, return_details = FALSE, verbose = FALSE)
```

Group rows that share any value across the chosen columns (Union–Find over rows).

| Argument | Description |
|---|---|
| `data` | A `data.frame`/`data.table`, or a list of equal-length columns. |
| `cols` | Column names, or regex patterns when `use_regex = TRUE`. `NULL` uses all columns. |
| `use_regex` | Treat `cols` as regex patterns matched against column names. |
| `incomparables` | Values that never link rows (empty, `"NA"`, etc.). |
| `case_sensitive` | Case-sensitive string comparison. |
| `min_group_size` | Groups smaller than this get id `0`. |
| `return_details` | Return a rich `group_id_result` object instead of a bare vector. |
| `verbose` | Print timing/progress. |

**Returns** an integer vector of group ids (one per row); or, with
`return_details = TRUE`, a `group_id_result` list with `group_ids`, `n_groups`,
`group_sizes`, and `value_map` (which shared values formed each group).

**Note** — an all-numeric, no-incomparables, case-sensitive call dispatches to a
specialised fast numeric path automatically.

## `print(x, ...)` for `group_id_result`

S3 print method summarising counts, settings, the first group ids, and the
shared values that created groups.

## `add_group_ids`

```{r}
add_group_ids(dt, cols, group_col = "group_id", use_regex = TRUE, ...)
```

`data.table` wrapper for `group_id()` — adds the group column **by reference**.
Extra `...` are passed through to `group_id()`.

**Returns** `dt` (invisibly).

## `set_group_id`

```{r}
set_group_id(dt, cols = "phone", var_output_name = "gid", incomparables = c(NA, ""))
```

Alternative grouping engine using the **bipartite edge reduction**
(`melt` → edges → `edge_components`).

| Argument | Description |
|---|---|
| `dt` | A `data.table` that **must contain an `id` column**. |
| `cols` | Regex matched against column names to choose the value columns. |
| `var_output_name` | Name of the group-id column to add. |
| `incomparables` | Values excluded from linking. |

**Returns** `dt`, modified by reference with the new group column.

---

## See also

- [Theory](theory.qmd) — how these algorithms work and why they are fast.
- [Examples](examples.qmd) — runnable demonstrations of every function.
- [Benchmarks](benchmarks.qmd) — scaling against `igraph`, `network`, `sna`.
